Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-11-13 article.Rnw
Improve background section

\section{Reparametrization}
\label{reparametrization}

The proposed reparametrization of COM-Poisson models is based on
mean approximation (\autoref{eqn:mean-approx}). We introduced new
parameter $\mu$, using the mean approximation
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The precision parameter is taken on the log scale to avoid restrictions
on the parameter space, $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ are the same to $\nu$, but in other scale. For
$\phi<0$ we have the overdispersed case and $\phi>0$ underdispersed,
relatives to Poisson distribution. $\phi=0$ represents the special case
that the COM-Poisson recovers to Poisson.

@@

#-------------------------------------------
# Mean and variance relationship
aux <- expand.grid(
    mu = seq(3, 30, length.out = 50),
    phi = seq(-1.5, 1.8, length.out = 50))

@@

In order to assess the accuracy of the moments approximations
(\autoref{eqn:mean-approx}) we presented the quadratic errors of moments
approximations in \autoref{fig:approx-plot}(a) and
\ref{fig:approx-plot}(b). Quadratic errors are obtained by
$[\mu - E(Y)]^2$ for expectation and by
$[ \mu \exp(-\phi) - \text{var}(Y)]^2$, where $E(Y)$ and var$(Y)$ are
numerically calculated using the moments definitio. The dotted lines
represent the border between the regions $\nu \leq 1$ and
$\lambda > 10^\nu$, in the $(\mu\,,\phi)$ scale.

<<approx-plot, fig.height=3, fig.width=8, fig.cap="Quadratic errors for expectation (a) and variance (b) approximation, dotted lines representing the restriction for good approximations by \\cite{Shmueli2005}. Mean-variance relationship (c), dotted line represents $E(Y)=\text{var}(Y)$.">>=

@@

myreg <- colorRampPalette(c("gray90",  "gray20"))
xy1 <- levelplot(emu ~ phi + mu, data = grid,
                 aspect = "fill",
                 col.regions = myreg(100),
                 xlab = expression(phi),
                 ylab = expression(mu),
                 sub = "(a)",
                 colorkey = list(space = "top"),
                 par.settings = ps2,
                 panel = function(x, y, z, ...) {
                     panel.levelplot(x, y, z, ...)
                     panel.curve(10 - (exp(x) - 1)/(2 * exp(x)),
                                 lty = 2)
                     panel.abline(v = 0, lty = 2)
                 })

xy2 <- levelplot(eva ~ phi + mu, data = grid,
                 aspect = "fill",
                 col.regions = myreg(100),
                 xlab = expression(phi),
                 ylab = expression(mu),
                 sub = "(b)",
                 colorkey = list(space = "top"),
                 par.settings = ps2,
                 panel = function(x, y, z, ...) {
                     panel.levelplot(x, y, z, ...)
                     panel.curve(10 - (exp(x) - 1)/(2 * exp(x)),
                                 lty = 2)
                     panel.abline(v = 0, lty = 2)
                 })

#-------------------------------------------
# Mean-variance relantionship
xy3 <- xyplot(var ~ mean,
              groups = phi,
              data = grid,
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(E(Y)),
              ylab = expression(var(Y)),
              sub = "(c)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = modifyList(
                  ps2, list(superpose.line = list(
                                col = myreg(length(unique(grid$phi))))
                            )
              ),
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(1*x, min(x), max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            x = unit(0.15, "npc"),
                            y = unit(0.83, "npc"))
              })

print(xy1, split = c(1, 1, 3, 1), more = TRUE)
print(xy2, split = c(2, 1, 3, 1), more = TRUE)
print(xy3, split = c(3, 1, 3, 1), more = FALSE)

@@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the maximum of the quadratic errors is
\Sexpr{max(grid[["emu"]])} in the range evaluated. For the variance
approximation, the maximum is \Sexpr{max(grid[["eva"]])} and this occurs
to negative values of $\phi$. Interestingly, the errors are bigger to
negative values of $\phi$ and appear to have no relation with $\mu$
against the regions gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).
\autoref{fig:approx-plot}(c) shows that the relationship
between mean and variance is linear with slope controlled by precision
parameter $\phi$, but not in the form $\mu \exp(-\phi)$. In terms of
moments, this leads to a specification indistinguishable from the
quasi-Poisson.

@@

leg_phi <- parse(
    text = paste("phi == \"",
                 formatC(unique(parg$phi), 1, format = "f"),
                 "\""))
barchart(py ~ y | factor(mu),
         groups = factor(phi),
         data = parg,
         horizontal = FALSE,
         layout = c(NA, 1),
         as.table = TRUE,
         axis = axis.grid,
         origin = 0,
         xlim = extendrange(y, f = 0.01),
         border = "transparent",
         scales = list(x = list(at = pretty(y))),
         ylab = expression(P(Y==y)),
         xlab = expression(y),
         par.settings = list(
             superpose.polygon = list(
                 col = c("gray40", "gray60", "gray10")),
             superpose.line = list(
                 col = c("gray40", "gray60", "gray10"),
                 lwd = 2)
         ),
         auto.key = list(
             columns = 3,
             rectangles = FALSE,
             lines = TRUE,
             text = leg_phi
         ),
         strip = strip.custom(
             strip.names = TRUE,
             var.name = expression(mu == ""),
             sep = ""))

@@

>>>>

\section{Reparametrized COM-Poisson regression model}
\label{reparametrization}

The proposed reparametrization of COM-Poisson models is based on
mean approximation (\autoref{eqn:mean-approx}). We introduced new
parameter $\mu$, using the this approximation
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but in another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have Poisson
distribution as the special case.

@@

#-------------------------------------------
# Mean and variance relationship
aux <- expand.grid(
    mu = seq(3, 30, length.out = 55),
    phi = seq(-1.5, 1.8, length.out = 50))

@@

In order to assess the accuracy of the approximations presented in
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
quadratic errors for the expectation (a) and variance (b).  The
quadratic errors were obtained by $[\mu - E(Y)]^2$ for the expectation
and by $[ \mu \exp(-\phi) - \text{var}(Y)]^2$ for the variance. In both
cases $E(Y)$ and var$(Y)$ were computed numerically. The dotted lines
represent the border between the regions $\nu \leq 1$ and
$\lambda > 10^\nu$, in the $\mu$ and $\phi$ scale.

<<approx-plot, fig.height=3, fig.width=6, out.width=".8\\textwidth", fig.cap="Quadratic errors for the approximation of the expectation (a) and variance (b). Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=

@@

myreg <- colorRampPalette(c("gray90",  "gray20"))
xy1 <- levelplot(emu ~ phi + mu, data = grid,
                 aspect = "fill",
                 col.regions = myreg,
                 xlab = expression(phi),
                 ylab = expression(mu),
                 sub = "(a)",
                 colorkey = list(space = "top"),
                 par.settings = ps2,
                 panel = function(x, y, z, ...) {
                     panel.levelplot(x, y, z, ...)
                     panel.curve(10 - ( exp(x) - 1)/(2 * exp(x)),
                                 lty = 2)
                     panel.abline(v = 0, lty = 2)
                 })

xy2 <- levelplot(eva ~ phi + mu, data = grid,
                 aspect = "fill",
                 col.regions = myreg,
                 xlab = expression(phi),
                 ylab = expression(mu),
                 sub = "(b)",
                 colorkey = list(space = "top"),
                 par.settings = ps2,
                 panel = function(x, y, z, ...) {
                     panel.levelplot(x, y, z, ...)
                     panel.curve((10 - ( exp(x) - 1)/
                                  (2 * exp(x)))/exp(x), lty = 2)
                     panel.abline(v = 0, lty = 2)
                 })

@@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the largest quadratic error is
\Sexpr{max(grid[["emu"]])} for the parameter values evaluated. For the
variance approximation, the largest quadratic error was
\Sexpr{max(grid[["eva"]])} and it occurs for negative values of
$\phi$. Interestingly, the errors are larger to negative values of
$\phi$ and present no clear relation with $\mu$ opposite to the regions
gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).

@@

leg_phi <- parse(
    text = paste("phi == \"",
                 formatC(unique(parg$phi), 1, format = "f"),
                 "\""))
barchart(py ~ y | factor(mu),
         groups = factor(phi),
         data = parg,
         horizontal = FALSE,
         layout = c(NA, 1),
         as.table = TRUE,
         axis = axis.grid,
         origin = 0,
         xlim = extendrange(y, f = 0.01),
         border = "transparent",
         # scales = list(x = list(at = pretty(y))),
         ylab = expression(P(Y==y)),
         xlab = expression(y),
         par.settings = list(
             superpose.polygon = list(
                 col = c("gray40", "gray60", "gray10")),
             superpose.line = list(
                 col = c("gray40", "gray60", "gray10"),
                 lwd = 2)
         ),
         auto.key = list(
             columns = 3,
             rectangles = FALSE,
             lines = TRUE,
             text = leg_phi
         ),
         strip = strip.custom(
             strip.names = TRUE,
             var.name = expression(mu == ""),
             sep = ""))

@@

In order to explore the flexibility of the COM-Poisson model to deal
with real count data, we compute indexes for dispersion (DI),
zero-inflation (ZI) and heavy-tail (HI), which are respectively given by
\begin{equation*}
\text{DI} = \frac{\text{var}(Y)}{\text{E}(Y)}, \quad
\text{ZI} = 1 + \frac{\log \Pr(Y = 0)}{\text{E}(Y)}
  \quad \text{and} \quad
\text{HT} = \frac{\Pr(Y=y+1)}{\Pr(Y=y)}\quad \text{for} \quad y \to
\infty.
\end{equation*}
These indexes are defined in relation to the Poisson distribution. Thus,
the dispersion index indicates overdispersion for $\text{DI} > 1$,
underdispersion for $\text{DI} < 1$ and equidispersion for
$\text{DI} = 1$. The zero-inflation index indicates zero-inflation for
$\text{ZI} > 0$, zero-deflation for $\text{ZI} < 0$ and no excess of
zeros for $\text{ZI} = 0$. Finally, heavy-tail index indicates a
heavy-tail distribution for $\text{HT} \to 1$ when $y \to \infty$.

These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions. Regarding the COM-Poisson$_\mu$
distribution, we presents in \autoref{fig:indexes-plot} the relationship
between mean and variance (a), the dispersion and zero-inflation indexes
(b--c) for differents values of $\mu$ and $\phi$ and heavy-tail index
(d) for $\mu=25$ and differents values of $y$ and $\phi$.

<<compute-indexes, include=FALSE, cache=TRUE>>=

# Dispersion index
grid <- transform(grid, dispersion_index = var / mean)

# Zero-inflation index
prob0 <- mapply(FUN = dcmp,
                mu = grid$mu,
                phi = grid$phi,
                MoreArgs = list(y = 0, sumto = 300),
                SIMPLIFY = FALSE)
prob0 <- unlist(prob0)
grid <- transform(grid, zero_index = 1 + log(prob0) / mean)

# Heavy tail-index
hi_fun <- Vectorize(FUN = function(x, mu, phi, sumto) {
    probs <- dcmp(y = c(x, x + 1), mu = mu, phi = phi, sumto = sumto)
    probs[2] / probs[1]
}, vectorize.args = "x")

x <- 40:150
his <- mapply(FUN = hi_fun,
              mu = grid$mu,
              phi = grid$phi,
              MoreArgs = list(x = x, sumto = 300),
              SIMPLIFY = FALSE)

aux1 <- lapply(his, function(k) data.frame("x" = x, "heavy_index" = k))
aux1 <- do.call(rbind, aux1)
aux2 <- grid[c("mu", "phi")][rep(1:nrow(grid), each = length(x)), ]
heavy_data <- cbind(aux1, aux2)

<<indexes-plot, fig.width=9.5, fig.height=4, fig.cap="Indexes for COM-Poisson distribution. Mean and variance relation ship (a), dispersion, zero-inflation and heavy-tail indexes (b--d) for different parameter values. Dotted lines represents the Poisson special case.">>=

#-----------------------------------------------------------------------
# Study indexes on the reparametrized COM-Poisson distribution
mygpar <- gpar(cex = 1.2)
psaux <- modifyList(
    ps0, list(superpose.line = list(
                  col = myreg(length(unique(grid$phi)))),
              layout.widths = list(
                  left.padding = -0.5)
              )
)

#-------------------------------------------
# Mean and variance relationship
xy1 <- xyplot(var ~ mu | "Mean-Variance",
              groups = phi,
              data = grid,
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(mu),
              ylab = "",
              # ylab = expression(var(Y)),
              scales = list(y = list(rot = 90)),
              sub = "(a)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = psaux,
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(1*x, min(x), max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            gp = mygpar,
                            x = unit(0.020, "npc"),
                            y = unit(0.875, "npc"))
              })

#-------------------------------------------
# Dispersion index (DI)
xy2 <- xyplot(var / mean ~ mu | "Dispersion index",
              groups = phi,
              data = grid,
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(mu),
              ylab = "",
              scales = list(y = list(rot = 90)),
              sub = "(b)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = psaux,
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(1 + 0*x, min(x), max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            gp = mygpar,
                            x = unit(0.020, "npc"),
                            y = unit(0.875, "npc"))
              })

#-------------------------------------------
# Zero-inflation index
xy3 <- xyplot(zero_index ~ mu | "Zero-inflation index",
              groups = phi,
              data = grid,
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(mu),
              ylab = "",
              scales = list(y = list(rot = 90)),
              sub = "(c)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = psaux,
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(0*x, min(x), max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            gp = mygpar,
                            x = unit(0.020, "npc"),
                            y = unit(0.875, "npc"))
              })

#-------------------------------------------
# Heavy tail-index
# Choose one specific mu of unique(grid$mu)
choosemu <- 25
xy4 <- xyplot(heavy_index ~ x | "Heavy-tail index",
              groups = phi,
              # data = heavy_data,
              data = subset(heavy_data, mu == choosemu),
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(y),
              ylab = "",
              scales = list(y = list(rot = 90)),
              sub = "(d)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = psaux,
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(choosemu / (x + 1), from = min(x),
                              to = max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            gp = mygpar,
                            x = unit(0.020, "npc"),
                            y = unit(0.875, "npc"))
              })

print(xy1, split = c(1, 1, 4, 1), more = TRUE)
print(xy2, split = c(2, 1, 4, 1), more = TRUE)
print(xy3, split = c(3, 1, 4, 1), more = TRUE)
print(xy4, split = c(4, 1, 4, 1), more = FALSE)

midi <- min(grid$dispersion_index)
madi <- max(grid$dispersion_index)

\autoref{fig:indexes-plot} shows that the indexes are slightly depend on
the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribuion is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle with a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicates the distribution is in general
light tailed distribution, i.e. $HT \to 0$ for $y \to \infty$.
